Thermal reactionome of a temperate ant species

### 2014-05-16 doc/ApTranscriptome_manuscript.Rmd

Colonies of the ant species *Aphaenogaster picea* were collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis* from Durham, North Carolina ( 36.037° N -78.874° W), designated as *ApVT* and *AcNC* respectively. The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice Bacon DeMarco, pers. communication). Colonies of these species were maintained in the lab for 6 months prior to experimentation.

@@

```{r expression_type_table, echo=FALSE, results='asis'}
pandoc.table(tt3, style="rmarkdown", split.tables = Inf, caption = "Number of thermally-responsive genes in each expression category for each species. *AcNC* in rows and *ApVT* in columns.")
```

@@

```{r expression_type_plot, echo=FALSE, fig.cap = "Mosaic plot that shows the proportion of thermally-responsive genes in each expression category for each species."}
mosaicplot(t(tt2), shade = FALSE, ylab = "AcNC", xlab = "ApVT", main = "", cex = 0.5)
```

@@

We quantified differences in the expression reaction norms of thermally-responsive genes between species as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta C$) and wiggle ($\Delta W$). Overall, we found mean expression was greater in *AcNC* (*t*~`r t.offset$parameter`~ = `r t.offset$statistic`, *P* < `r max(t.offset$p.value, 0.001)`),

but changes in the shape due to slope , curvature (*t* = 15.71, *P* < 0.001) and wiggle (*t* = -12.45, *P* < 0.001) were all highly significant. On average, slope (95% CI 0.41 - 0.48) and curvature (95% CI 0.25 - 0.33) were greater in *A. picea* than *A. carolinensis*, indicating a steeper response to temperature.

@@

Of the total differences ($\Delta T$) in reaction norms between species for each gene, about 8% (95% CI 1 - 32%) was due to shifts in mean, 47% (95% CI 14 - 92%) due to shifts in slope, 39% (95% CI 0 - 67%) due to shifts in curvature and 0% (95% CI 0 - 87%) due to shifts in wiggle.

Plasticity is a well-characterized mechanism by which organism tolerate environmental change. Given our sampling of a warmer (*A. carolinensis*) and colder (*A. picea*) climate species, we hypothesized that genes which are plastically-expressed under heat shock in the colder species would have greater expression near optimal temperatures of 19°C in the warmer-climate species, and vice versa. Partially consistent with our prediction, we found that optimal-temperature expression was greater in the more southern *A. carolinensis* species for *High* genes in *A. picea* (Wilcoxon V = `r w1$statistic`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r w1$conf.int[1:2]`). However, for genes with *Low* expression in *A. carolinensis* we also found that optimal-temperature expression was greater in *A. carolinensis* than *A. picea* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`). Moreover, we also found that for genes with *Intermediate* expression in either colony, optimal-temperature expression was greater in *A. carolinensis*. The overall conclusion from these analyses is that for thermally-responsive genes, the more southern species, *A. carolinensis* has greater optimal-temperature expression.

We tested for a transcriptome-wide pattern of thermal tolerance and sensitivity by examining the thermal range for the *Intermediate* expressed genes. Contrary to our expectations based on critical maximum and minimal temperatures, *A. carolinensis* had a transcriptome-wide pattern of broader thermal range (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64). In contrast, there was no difference in thermal sensitivity as measured by the standard deviation of expression for *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76).

We also explored the extent to which the temperature at which thermally-responsive genes are activated differs between species. For genes that are upregulated at *High* temperatures, including both *High* and *Bimodal* genes, we found that ... In contrast, for genes that are upregulated at *Low* temperatures, including both *Low* and *Bimodal* genes, we found that...

@@

>>>>

Colonies of the ant species *Aphaenogaster picea* were collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis* from Durham, North Carolina ( 36.037° N -78.874° W). The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice Bacon DeMarco, pers. communication). Colonies of these species were maintained in the lab for 6 months prior to experimentation.

@@

```{r expression_type_table, echo=FALSE, results='asis'}
pandoc.table(tt3, style="rmarkdown", split.tables = Inf, caption = "Number of thermally-responsive genes in each expression category for each species. *A. carolinensis* in rows and *A. picea* in columns.")
```

@@

```{r expression_type_plot, echo=FALSE, fig.cap = "Mosaic plot that shows the proportion of thermally-responsive genes in each expression category for each species."}
mosaicplot(t(tt2), shade = FALSE, ylab = "A. carolinensis", xlab = "A. picea", main = "", cex = 0.5)
```

@@

We quantified differences in the expression reaction norms of thermally-responsive genes between species as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta C$) and wiggle ($\Delta W$). Mean expression (*t*~`r t.mean$parameter`~ = `r t.mean$statistic`, *P* < `r max(t.mean$p.value, 0.001)`) was about `r round(exp(t.mean$estimate)` TPM greater in *A. carolinensis*. Mean slope (*t*~`r t.slope$parameter`~ = `r t.slope$statistic`, *P* < `r max(t.slope$p.value, 0.001)`) was about `r round(exp(t.mean$slope)` $\Delta$TPM/degree greater in *A. picea*. Curvature (*t*~`r t.curvature$parameter`~ = `r t.curvature$statistic`, *P* < `r max(t.curvature$p.value, 0.001)`) and wiggle (*t*~`r t.wiggle$parameter`~ = `r t.wiggle$statistic`, *P* < `r max(t.wiggle$p.value, 0.001)`) did not differ between species. Of the total differences ($\Delta T$) in reaction norms between species for each gene, about 8% (95% CI 1 - 32%) was due to shifts in mean, 47% (95% CI 14 - 92%) due to shifts in slope, 39% (95% CI 0 - 67%) due to shifts in curvature and 0% (95% CI 0 - 87%) due to shifts in wiggle.

@@

By sampling a warmer (*A. carolinensis*) and colder (*A. picea*) climate species, we were able to test if the same genes that are up-regulated in response to thermal stress have increased basal expression in the other species. Partially consistent with this prediction, basal expression was greater in the more southern *A. carolinensis* species for *High* genes in *A. picea* (Wilcoxon V = `r w1$statistic`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r w1$conf.int[1:2]`). However, for genes with *Low* expression in *A. carolinensis* basal expression was also greater in *A. carolinensis* than *A. picea* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`). Moreover, basal expression was also greater in *A. carolinensis* for genes with *Intermediate* expression in either colony. The overall conclusion from these analyses is that for thermally-responsive genes, the more southern species, *A. carolinensis* has greater optimal-temperature expression.

We tested for a transcriptome-wide pattern of thermal tolerance and sensitivity by examining the thermal range for the *Intermediate* and *Bimodal* expressed genes. Contrary to our expectations based on critical maximum and minimal temperatures, *A. carolinensis* had a broader thermal range of stable gene expression (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64). In contrast, there was no difference in thermal sensitivity among species for the *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76).

We also explored the extent to which the temperature at which thermally-responsive genes are activated differs between species. For genes that are upregulated at *High* temperatures, we found that *A. carolinensis* genes were activated on average 1°C hotter than *A. picea*. In contrast, for genes that are upregulated at *Low* temperaturese we found that...

@@

overall pattern *A. carolinensis*

- more genes thermally-responsive
- tolerates thermal stress by shutting down biological processes

overall pattern *A. picea*

- more *Low* genes
